working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect

### 2018-12-19 Manuscript_draft/Manuscript_draft.tex
changes

In most ecosystems, plant species normally coexist and share their
floral visitors with other species Bascompte et al. (2003). From the
plants' perspective, this pollinator sharing can be positive due to
facilitation Carvalheiro et al. (2014) or negative due to competition at
the pre-pollination stage Pauw (2013). An increasing number of visits
often correlates with higher chances of fertilization Engel and Irwin
(2003). However this is not always the case, among these possible flower
visitors there are also nectar robbers and pollen thieves Inouye (1980).
Receiving both sufficient quantity and quality deposited on the stigma
is thus highly relevant to the pollination success of the plant Aizen
and Harder (2007).

@@

By visiting many plant species, many pollinators are responsible for
conspecific pollen loss and the transport of foreign pollen, both of
which can have important detrimental effects on species fitness Morales
and Traveset (2008); Ashman and Arceo-Gómez (2013); Arceo-Gómez and
Ashman (2016). Foreign pollen arrival can play an important role in
plant species fitness but outcomes are variable and appear to be context
dependent as there is not always a decrease in fitness Morales and
Traveset (2008). Some of this variation is likely due to the enormous
variability of foreign pollen transferred across systems ranging from
between 0 and 75 percent but most studies report ranges of
heterospecific pollen between 0 and 20 percent of the total pollen load
Bartomeus et al. (2008) Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Fang and Huang (2013), yet even these relatively low
amounts of heterospecific pollen transferred can decrease fitness
greatly Thomson et al. (1982). While we now have some understanding of
the impacts of heterospecific pollen quantity, we have less
understanding of other factors that could be driving the variation in
impacts upon fitness. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework that identifies a need to understand how plant
traits might mediate heterospecific pollen effect, whereby mating system
and pollen size were predicted to potentially mediate the impact of
foreign pollen transfer on plant fitness. This concept is supported by
other studies, such as Tong and Huang (2016) that demonstrate an
asymmetrical effect in 6 species of \emph{Pedicularis} whereby the
pollen of long styled species was able to grow the full length of the
style on short styled species but not vice versa. While this suggests
that the impacts of heterospecific pollen may differ among pollen donor
and recipient, few studies have been conducted to ascertain whether this
pattern is in fact a trend or to identify the extent to which other
plant traits are critical to heterospecific pollen impacts.

@@

Given the large variability in mating systems across populations
Whitehead et al. (2018), it is difficult to determine potential impacts
upon HP transfer yet incompatibility system is another plant trait that
appears to play an important role in foreign pollen effect whereby
species that are self incompatible have stronger barriers to
heterospecific pollen than self-compatible species Ashman and
Arceo-Gómez (2013). The type of incompatibility, (i.e.~whether
sporophytic or gametophytic) is related to the location of pollen
recognition; sporophytic incompatibility relates to signaling at the
stigma surface while gametophytic occurs within the style Barrett
(1988). This later acting pollen recognition mechanism is associated
with greater negative effect than sporophytic recognistion Ashman and
Arceo-Gómez (2013). Nonetheless, an effect of foreign pollen is a bit
obscured by the variability within species, however species that are
strong selfers or strong outcrossers have less variablity in mating
systems and predictions of effect could be more realistic (see figure 1
from Whitehead et al. (2018)). Moreover, other traits such as number of
pollen grains per flower and number of ovules have been traditionally
associated with the type of incompatibility system where species with
higher pollen ovule ratios are predicted to be xenogamous and species
with low pollen ovule ratios autogamous (REF). Selfer species are known
to have a reduction of herkogamy (REF) and less pollen production per
ovule (REF) which can be interpretated as a reduction of pollen exported
into the community. Other morphological traits, like stigma size can be
determinant for the total pollen quatity that a stigma can receive and
therefore related to do that pollen size would also play an important
role. Example with pollen here.

@@

Species with similar traits are more closely related XXXXXXXXX. (Refs?
Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang
(2016) ). Several studies predict that the impact of HP transfer is
likely to be greater for closely related species (Ashman and Arceo-Gómez
(2013)). Few studies however, have focused on the impacts of
heterospecific pollen of distantly related species Thomson et al. (1982)
Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects
and most stigmas have been found to carry multiple species of foreign
pollen with little attention to degree of relatedness (Arceo-Gómez and
Ashman (2016); Fang and Huang (2013) ; also cite studies from pollen
transfer networks here such as\ldots{} ). Further, a majority of plant
species are generalist and thus receive visits from multiple different
pollinators. Given these are generally the ones that receive greater
loads of heterospecific pollen Fang and Huang (2013) and unrelated
species are more likely to coexist with other species due to less niche
overlap (Ref), understanding the role of foreign pollen from distantly
related species thus deserves greater attention in understanding
coexistence blah blahXXXXX refs.. Notwithstanding, the effect of
heterospecific pollen of far and close related species at community
level remains to be explored beyond single pairwise interactions.

@@

The study was conducted in a glasshouse at University of New England
(Armidale, Australia) from November 2017 to March 2018. Rooms were
temperature controlled depending on the requirements of the species with
day and night temperature differences. The species selected
(\textbf{Table 1}) belonged to three different families, Solanaceae,
Brassicaceae and Convolvulaceae. The criteria of species/family
selection was based on close/distant related species (see phylogenetic
tree for relatedness fig 1), heterogeneous traits, low structural flower
complexity and fast life cycle. For the purpose of the experiment all
the species where considered as pollen recipient and as pollen donor
(see interaction matrix, fig 2). Species were watered once or twice per
day and fertilized weekly (NPK 23: 3.95: 14).

@@

Foreign pollen effect was studied through two different treatments, one
with 50\% conspecific pollen and 50\% heterospecific pollen and a second
one with 100\% foreign pollen (N=10). Therefore, 180 different
combinations were perform with N=10. Seed set was the proxy of effect
for all our treatments. Moreover, hand cross pollination, hand self
pollination, apomixis (bagged emasculated flowers) and natural selfing
were tested for each species (N=10). Flowers were emasculated the day
prior anthesis and hand pollinated next day with a toothpick.
Hand-pollination was conducted with 3-4 gentle touches on the stigma
surface. The mixes of pollen were realized on an eppendorf based on the
pollen counts maded with Neubaeur chamber (each anther was counted 4
times for 20 different anthers per species). In order to confirm that
the treatments applied were 50-50 percent pollen, for each focal species
the total stigmatic load of pollen was counted from one donor of each
family (N=3).

@@

The traits measured for each species were pollen per anther, number of
ovules, stigma width and length and stigmatic area, style width and
length, ovary width and length. Moreover stigma type was tested. Pollen
was counted for 20 anthers of each species with 4 replicates per sample
with an hemocytometer. Previously anthers were squashed on a known
solution with the pippete tip and homogeneize with a vortex for 30
seconds. Ovule number was counted with the help of an stereomicroscope
and a small grid over a petri dish from 15 randomly selected flowers.
The different morphometrical traits were measured with a digital
stereomicrospe. Levels of self incompatibility were estimated by
dividing the the fruit set of hand self pollination by hand cross
pollination Lloyd and Schoen (1992).

@@

We used the statistical language \texttt{R} (R Core Team 2018) for all
our analyses. Differences of seed set between treatments and hand cross
pollination for each species was tested through mixed linear models. For
the following analysis we scaled the values of seed production for all
the species with mean 0 and sd of 1. To test the effect of
heterospecific pollen, we substracted to the seed set of hand cross
pollination the seed set of heterospecific pollen treatments. In order
to see correlations between heterospecific pollen effect and traits we
performed Mantel test between the matrix of heterospecific pollen effect
and the distance matrix of each trait (euclidean distances). Moreover,
Mantel test was also conducted between heterospecific pollen effect and
the square root of the matrix of phylogenetic distance due to
improvement in the statistical power (Letten \& Cornwell 2014). We
explored also the the relations between traits and heterospecific pollen
effect through generalized mixed models where the response variable was
heterospecific pollen effect, the independent variable the different
traits and the random effects the different treatments per species.
Moreover, pairwise evolutive distances were calculated with MEGA7 for
two kinds of markers: 1) Internal transcribed spacer (ITS) and 2)
ribulose-bisphosphate carboxylase (RBCL). The sequences of interest were
downloaded from NCBI GenBank and the phylogenetic tree constructed by
maximum likelihood with MEGA7.

@@

Results of hand cross pollination, self hand pollination, natural
selfing and apomixis are presented in \textbf{Table 2}. Heterospecific
pollen reduced seet set signifcatively with the 50-50\% heterospecific
pollen treatments for 65\% of the pairwise interactions
p\textless{}0.05. Across families we found a very similar effect but
when species where look at species level they respond differently even
within the same family, for instance for two species of the Brassicaceae
family \emph{Brassica oleracea} and \emph{Eruca versicaria} we found
very contrasting effects of foreign pollen where for the first one, all
donors reduce seed set significatively and for the second, just two
species did out of nine. The 100\% foreign pollen treatments barely
produced seeds or fruits and just for \emph{Sinapis alba} we did not
find significant differences between the hand cross pollination and one
treatment with pollen from a confamilial. Solanaceae species with berry
fruit type developed small fruits or even normal fruits in some cases.
\emph{S. lycopersicum} seems to produced small fruits (35\% of the
treatments) independently of pollen and pollen donor due to also
apomictic treatments did, never normal size. \emph{C. annuum} produced
some fruits (9\%) of both small and normal size and finally \emph{S.
melongena} produced seedless normal fruits with just confamilial pollen
(3\%), for both species seems that fruit formation was induced by pollen
on the stigma because of lack of fruit production with treatments that
tested for apomixis.

\textbf{Table 2}. Perecentage of seeds produced per ovule for the ten
species used in the experiment. The treatments presented are hand cross
pollination, hand self pollination, natural selfing and apomixis
(emasculated flowers).

@@

Mantel test indicates that a possible correlation exist between
heterospecific pollen effect and the evolutive distances, for ITS and
RBCL markers we had r coefficients of 0.29 and 0.25 respectively
p\textless{}0.05. Moreover, Mantel test indicates that also a possible
correlation between stigma width and stigma type exist. Trait
correlations were also explored with GLMM

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-3-1.pdf}
\caption{The effect of heterospecific pollen (scaled see set) is
represented in function of the compatibility system (self/cross*100) for
the the different species. Each coulored dot represents the interaction
of a focal species with a different pollen donor.}
\end{figure}

@@

@@

\hypertarget{ref-bascompte2003}{}
Bascompte, J., P. Jordano, C. J. Melián, and J. M. Olesen. 2003. The
nested assembly of plant--animal mutualistic networks. Proceedings of
the National Academy of Sciences 100:9383--9387.

@@

@@

@@

>>>>

In most ecosystems, plant species normally coexist and share their
floral visitors with other species Waser et al. (1996). From the plants'
perspective, pollinator sharing can be positive for some plants
Carvalheiro et al. (2014) or negative for others Pauw (2013), depending
on the facilitation gradient. An increasing number of visits often
correlates with higher chances of fertilization Engel and Irwin (2003).
However this is not always the case, among these possible flower
visitors there are also nectar robbers and pollen thieves Inouye (1980);
Magrach et al. (2017). Receiving both sufficient quantity and quality
deposited on the stigma is thus highly relevant to the pollination
success of the plant Aizen and Harder (2007).

@@

By visiting many plant species, many pollinators are responsible for
conspecific pollen loss and the transport of foreign pollen, both of
which can have important detrimental effects on species fitness Morales
and Traveset (2008); Ashman and Arceo-Gómez (2013); Arceo-Gómez and
Ashman (2016). Foreign pollen arrival can play an important role in
plant species fitness but outcomes are variable and appear to be context
dependent as there is not always a decrease in fitness Morales and
Traveset (2008). Some of this variation is likely due to the enormous
variability of foreign pollen transferred across systems ranging from 0
to 75 percent. However, most studies report ranges of heterospecific
pollen between 0 and 20 percent of the total pollen load Bartomeus et
al. (2008) Montgomery and Rathcke (2012); Ashman and Arceo-Gómez (2013);
Fang and Huang (2013), yet even these relatively low amounts of
heterospecific pollen transferred can decrease fitness greatly Thomson
et al. (1982). While we now have some understanding of the impacts of
heterospecific pollen quantity, we have less understanding of other
factors that could be driving the variation in impacts upon fitness.
Ashman and Arceo-Gómez (2013) postulated the first predictive framework
that identifies a need to understand how plant traits might mediate
heterospecific pollen effect, whereby mating system and pollen size were
predicted to potentially mediate the impact of foreign pollen transfer
on plant fitness. This concept is supported by specific case studies,
such as Tong and Huang (2016) that demonstrate an asymmetrical effect in
6 species of \emph{Pedicularis} whereby the pollen of long styled
species was able to grow the full length of the style on short styled
species but not vice versa. While this suggests that the impacts of
heterospecific pollen may differ among pollen donor and recipient, few
studies have been conducted to ascertain whether this pattern is in fact
a general trend or to identify the extent to which other plant traits
are critical to heterospecific pollen impacts.

@@

Incompatibility system is another plant trait that appears to play an
important role in foreign pollen effect. Species that are self
incompatible have stronger barriers to heterospecific pollen than
self-compatible species Ashman and Arceo-Gómez (2013). However, note the
large variability in mating systems across populations Whitehead et al.
(2018) \href{}{\ldots{}}. The type of incompatibility, (i.e.~whether
sporophytic or gametophytic) is related to the location of pollen
recognition; sporophytic incompatibility relates to signaling at the
stigma surface while gametophytic occurs within the style Barrett
(1988). This later acting pollen recognition mechanism is associated
with greater negative effect than sporophytic recognistion Ashman and
Arceo-Gómez (2013). \href{}{if you introduce this,I eexpect you will use
it in your analysis, remove} (Nonetheless, an effect of foreign pollen
is a bit obscured by the variability within species, however species
that are strong selfers or strong outcrossers have less variablity in
mating systems and predictions of effect could be more realistic (see
figure 1 from Whitehead et al. 2018)- IB this is a side problem for you,
mention just once, and maybe in discusion. Now it has a lot of
weigth.)(). Moreover, other traits such as number of pollen grains per
flower and number of ovules have been traditionally associated with the
type of incompatibility system where species with higher pollen ovule
ratios are predicted to be xenogamous and species with low pollen ovule
ratios autogamous (REF) . Selfer species are known to have a reduction
of herkogamy (REF) and less pollen production per ovule (REF) which can
be interpretated as a reduction of pollen exported into the community
\href{}{so, pollen and ovules are important by themselfs, or only
because correlated? Unclear what your point is}. Other morphological
traits, like stigma size can be determinant for the total pollen quatity
that a stigma can receive and therefore related to do that pollen size
would also play an important role. Example with pollen here.

@@

Species with similar traits are more closely related XXXXXXXXX \href{}{I
would say no. species closely related usually have similar traits,
specally if those are phylogenetially conserved}. (Refs? Brown and
Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) ).
Several studies predict that the impact of HP transfer is likely to be
greater for closely related species (Ashman and Arceo-Gómez (2013)). Few
studies however, have focused on the impacts of heterospecific pollen of
distantly related species Thomson et al. (1982) Galen and Gregory (1989)
Neiland and Wilcock (1999). Yet, most insects and most stigmas have been
found to carry multiple species of foreign pollen with little attention
to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang
(2013) ; also cite studies from pollen transfer networks here such
as\ldots{} ).

\href{}{here you change topic, new paragraph?} Further, a majority of
plant species are generalist and thus receive visits from multiple
different pollinators. Given these are generally the ones that receive
greater loads of heterospecific pollen Fang and Huang (2013) and
unrelated species are more likely to coexist with other species due to
less niche overlap (Ref), understanding the role of foreign pollen from
distantly related species thus deserves greater attention in
understanding coexistence blah blahXXXXX refs.. Notwithstanding, the
effect of heterospecific pollen of far and close related species at
community level remains to be explored beyond single pairwise
interactions.

@@

The study was conducted in a glasshouse at University of New England
(Armidale, Australia) from November 2017 to March 2018. Rooms were
temperature controlled depending on the requirements of the species with
day and night temperature differences. The species selected
(\textbf{Table 1}) belonged to three different families, Solanaceae,
Brassicaceae and Convolvulaceae. The criteria of species/family
selection was based on close/distant related species (see phylogenetic
tree for relatedness fig 1)\href{}{I would explain more the bauty of our
nested dessign to ensure close and far distance simultaneously},
heterogeneous traits, low structural flower complexity and fast life
cycle. For the purpose of the experiment all the species where
considered as pollen recipient and as pollen donor (see interaction
matrix, fig 2). Species were watered once or twice per day and
fertilized weekly (NPK 23: 3.95: 14).

@@

Foreign pollen effect was studied through two different treatments, one
with 50\% conspecific pollen and 50\% heterospecific pollen and a second
one with 100\% foreign pollen (N=10) \href{}{this second I don't get,
maybe explain it's ultility.}. Therefore, 180 different combinations
were performed with N=10 per combination. Seed set was the proxy of
effect for all our treatments. Moreover, hand cross pollination (between
individuals of the same species), hand self pollination, apomixis
(bagged emasculated flowers) and natural selfing were tested for each
species (N=10). For the treatments with foreign pollen and hand cross
pollination, flowers were emasculated the day prior anthesis and hand
pollinated next day with a toothpick. Hand-pollination was conducted
with 3-4 gentle touches on the stigma surface. \href{}{The mixes of
pollen were realized on an eppendorf based on the pollen counts made
with Neubaeur chamber (each anther was counted 4 times for 20 different
anthers per species)-IB explain better and give a bit more of detail}.
In order to confirm that the treatments applied were 50-50 percent
pollen, for each focal species the total stigmatic load of pollen was
counted from one donor of each family (N=3).

@@

The traits measured for each species were pollen per anther, number of
ovules, stigma width and length and stigmatic area, style width and
length, ovary width and length. Moreover stigma type \href{}{explain}
was tested. Pollen was counted for 20 anthers of each species with 4
replicates per sample with an hemocytometer. Previously, anthers were
squashed on a known solution with the pippete tip and homogeneize with a
vortex for 30 seconds. Ovule number was counted with the help of an
stereomicroscope and a small grid over a petri dish from 15 randomly
selected flowers. The different morphometrical traits were measured with
a digital stereomicrospe. Levels of self incompatibility were estimated
by dividing the fruit set of hand self pollination by hand cross
pollination Lloyd and Schoen (1992).

@@

We used the statistical language \texttt{R} (R Core Team 2018) for all
our analyses. Differences of seed set between treatments and hand cross
pollination for each species was tested through mixed linear models. For
the following analysis we scaled the values of seed production for all
the species with mean 0 and sd of 1. To test the effect of
heterospecific pollen, we substracted to the seed set of hand cross
pollination the seed set of heterospecific pollen treatments. In order
to see correlations between heterospecific pollen effect and traits we
performed Mantel test between the matrix of heterospecific pollen effect
and the distance matrix of each trait (euclidean distances). Moreover,
Mantel test was also conducted between heterospecific pollen effect and
the square root of the matrix of phylogenetic distance due to
improvement in the statistical power (Letten \& Cornwell 2014).
\href{}{all is here, but I would break it by questions and give a bit
more detail, to avoid overwhelm the reader} We explored also the
relations between traits and heterospecific pollen effect through
generalized mixed models where the response variable was heterospecific
pollen effect, the independent variable the different traits and the
random effects the different treatments per species {[}Here I think you
should think if this controls for the non independency of donors and
recipients. I think not. Maybe look onto matrix regresions?). Moreover,
pairwise evolutive distances were calculated with MEGA7 for two kinds of
markers: 1) Internal transcribed spacer (ITS) and 2)
ribulose-bisphosphate carboxylase (RBCL). The sequences of interest were
downloaded from NCBI GenBank and the phylogenetic tree constructed by
maximum likelihood with MEGA7.\href{}{Make a section on how you
contrsucted phylogeny}.

\href{}{I would explain three test. 0) treatment effects with GLM's, 1)
Mantels: relative effects, 2) GLM's or matrix models: Absolute effects
and explain them in three independent paragraphs including a rationale
of why}

@@

Results of hand cross pollination, self hand pollination, natural
selfing and apomixis are presented in \textbf{Table 2}. Heterospecific
pollen reduced seet set signifcatively with the 50-50\% heterospecific
pollen treatments for 65\% of the pairwise interactions
p\textless{}0.05. Across families we found a very similar effect but
when species where look at species level they respond differently even
within the same family \href{}{rephrase and maybe test statistically?},
for instance for two species of the Brassicaceae family \emph{Brassica
oleracea} and \emph{Eruca versicaria} we found very contrasting effects
of foreign pollen where for the first one, all donors reduce seed set
significatively and for the second, just two species did out of nine.
The 100\% foreign pollen treatments barely produced seeds or fruits
\href{}{and just for \emph{Sinapis alba} we did not find significant
differences between the hand cross pollination and one treatment with
pollen from a confamilial- IB Unclear}. Solanaceae species with berry
fruit type developed small fruits or even normal fruits in some cases
\href{}{under which treatment}. \emph{S. lycopersicum} seems to produced
small fruits (35\% of the treatments) independently of pollen and pollen
donor due to also apomictic treatments did, never normal size. \emph{C.
annuum} produced some fruits (9\%) of both small and normal size and
finally \emph{S. melongena} produced seedless normal fruits with just
confamilial pollen (3\%), for both species seems that fruit formation
was induced by pollen on the stigma because of lack of fruit production
with treatments that tested for apomixis.\href{}{clarify this
descriptive statistics part- Also a figure with a summary of the
treatments effect would be cool, or at least in the appendices}

\textbf{Table 2}. Perecentage of seeds produced per ovule for the ten
species used in the experiment. The treatments presented are hand cross
pollination, hand self pollination, natural selfing and apomixis
(emasculated flowers). \href{}{turn into a figure somehow?}

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-3-1.pdf}
\caption{Barplot with the different treatments that provide information
of the reproductive biology of the ten species. The y axis is the
proportion of ovules converted to seed in percentage. The different,
treatments which are presented in the legend are, hand cross
pollination, hand self pollination, natural selfing and apomixis. More
information about this treatments can be found in Methods and
Appendices.}
\end{figure}

Mantel test indicates that a possible?? \href{}{It exists!} correlation
exist between heterospecific pollen effect and the evolutive relative
distances, for ITS and RBCL markers we had r coefficients of 0.29 and
0.25 respectively p\textless{}0.05 \href{}{think on a figure - maybe
using NMDS}. Moreover, Mantel test indicates that also a possible??
correlation between stigma width and stigma type exist (stats??). Trait
correlations were also explored with GLMM

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-4-1.pdf}
\caption{The effect of heterospecific pollen (scaled see set) is
represented in function of the compatibility system (self/cross*100) for
the the different species. Each coulored dot represents the interaction
of a focal species with a different pollen donor.}
\end{figure}

@@

IB: Think also on using tree analysis to test if hp effect depends on
complex trait combinations. Tree analysis are great when two different
strategies lead to the same outcome. This would never been pick up by
GLMs. The r package is party\{\} . You can see an example applied to
birds is Sol et al 2010 Science. Ask me if you want more details or code
examples.

@@

@@

\hypertarget{ref-magrach2017}{}
Magrach, A., J. P. González-Varo, M. Boiffier, M. Vilà, and I.
Bartomeus. 2017. Honeybee spillover reshuffles pollinator diets and
affects plant reproductive success. Nature ecology \& evolution 1:1299.

@@

\hypertarget{ref-pauw2013}{}
Pauw, A. 2013. Can pollination niches facilitate plant coexistence?
Trends in ecology \& evolution 28:30--37.

@@

\hypertarget{ref-waser1996}{}
Waser, N. M., L. Chittka, M. V. Price, N. M. Williams, and J. Ollerton.
1996. Generalization in pollination systems, and why it matters. Ecology
77:1043--1060.
